Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-13 priority-queues-arxiv.tex
arxiv

We consider priority queues whose interface contains two methods $\textit{put}$ and $\textit{rm}$ for adding and respectively, removing a value. Each value is assigned with a priority when being added to the data structure (by calling $\textit{put}$) and the remove method $\textit{rm}$ removes a value with a minimal priority. For generality, we assume that the set of priorities is partially-ordered. Incomparable priorities can be removed in any order. When multiple values are assigned with the same priority, $\textit{rm}$ returns the least recent value. Also, when the set of values stored in the priority queue is empty, $\textit{rm}$ returns the distinguished value $\textit{empty}$. In this section, we formalize (concurrent) executions and implementations, introduce a set of properties satisfied by all the implementations we are aware of, and recall the standard correctness criterion for concurrent implementations of ADTs known as \emph{linearizability}~\cite{journals/toplas/HerlihyW90}.
%
%recall the notion of \emph{linearizability}, the standard correctness criterion for concurrent implementations of priority queues (
%
%Concurrent implementations of the priority queue allow the methods $\textit{put}$ and $\textit{rm}$ to be called concurrently from different threads. However, every method invocation should give the illusion that it takes place instantaneously at some point between its invocation and its return. We formalize (concurrent) executions and implementations in Section~\ref{ssec:exec}, Section~\ref{ssec:semantic_prop} introduces a set of properties satisfied by all the  implementations we are aware of, and Section~\ref{ssec:lin} defines the standard correctness criterion for concurrent implementations of ADTs known as \emph{linearizability}~\cite{journals/toplas/HerlihyW90}.

@@

We define $\seqPQ$, the set of sequential priority queue executions, semantically via a labelled transition system (LTS, for short). An LTS is a tuple $A=(Q,\Sigma,\rightarrow,q_0)$, where $Q$ is a set of states, $\Sigma$ is an alphabet of transition labels, $\rightarrow\subseteq Q\times\Sigma\times Q$ is a transition relation, and $q_0$ is the initial state. We model the priority queue as an LTS $\textit{PQ}$ where states are mappings associating priorities in $\mathbb{P}$ with sequences of values in $\mathbb{D}$, representing a snapshot of the priority queue (for each priority, the values are ordered as they were inserted), and the transition labels are $\textit{put}(a,p)$ and $\textit{rm}(a)$. Each transition modifies the state as expected. For example, $q_1 \xrightarrow{\textit{rm}(\textit{empty})} q_2$ if $q_1 = q_2$, and $q_1$ and $q_2$ map each priority to the empty sequence $\epsilon$. Then, $\seqPQ$ is the set of traces (words) accepted by $\textit{PQ}$. %}{\color {red}$\seqPQ$ is the set of traces of $\textit{PQ}$}. {\color {red}The detailed definition of $\textit{PQ}$ can be found in the long version \cite{CONCUR2017Ahmed}.}
The detailed definition of $\textit{PQ}$ can be found in Appendix \ref{sec:appendix definition of seqPQ and proof of Lemma EQP rules and semantics}.

An implementation $\mathcal{I}$ is a set of executions. Implementations represent libraries whose methods are called by external programs. In the remainder of this work, we consider only \emph{completed} executions, where each call action has a corresponding return action. This simplification is sound when the method invocations can always make progress in isolation.
%formally, for any execution $e$ with pending operations, there exists an execution $e'$ obtained by extending $e$ only with the return actions of the pending operations of $e$.
%Intuitively this means that methods can always return without any help from outside threads, avoiding deadlock.

%a finite set $\mathbb{M}$ of methods, and an infinite set $\mathbb{O}$ of operation (identifiers). Given $m \in \mathbb{M}$, $x \in \mathbb{D}$, and $o \in \mathbb{O}$, a call action $\textit{call}_o (m,x)$ represents an invocation to method $m$ with argument $x$, while a return action $\textit{ret}_o (m,x)$ represents a response from method $m$ with return value $x$. \footnote{Call actions with more than one arguments are similarly defined.} Here $o$ is used to match return actions to their call actions: $\textit{cal}_o (m,x)$ matches $\textit{ret}_{o'} (m',x')$, if $o=o' \wedge m=m'$. A sequential execution is a sequence of call and return actions, while each call action is immediately followed by its matching return action. Let $\cdot$ be the concatenation of sequences. To ease the reading, given a sequential execution $e=\textit{call}_{o_1}(m_1,a_1) \cdot \textit{rm}_{o_1}(m_1,b_1) \cdot \ldots \cdot \textit{call}_{o_n}(m_n,a_n) \cdot \textit{rm}_{o_n}(m_n,b_n)$, when the context is clear, we can write it as $e=m_1(a_1,b_1) \cdot \ldots \cdot m_n(a_n,b_n)$. We write $m(a,b)$ as $m(a)$ or $m(b)$, when there is no return value or there is no argument, respectively.

%A priority queue with partially-ordered priorities (priority queue for short) contains two method: $\textit{put}$ and $\textit{rm}$. A $\textit{put}$ method has two arguments, while the first argument is an item and the second argument is its priority. A $\textit{put}$ method is used to put an item into the priority queue with certain priority. Here the item is chosen from a specific (possibly infinite) data domain $\mathbb{D}$ and priority is chosen from a (possibly infinite) set $\mathbb{P}$. Moreover, there is a strict partial-order $\prec$ among elements in $\mathbb{P}$. A $\textit{rm}$ method intends to remove the item with minimal priority (w.r.t $\prec$) in priority queue and then returns it. It works as follows:

%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] If the priority queue is empty, then $\textit{rm}$ returns $\textit{empty}$.
%
%\item[-] Else, $\textit{rm}$ choose one of minimal priority of items in priority queue, and returns the earliest putted item of this priority. Note that if there are more than one incomparable and minimal candidate priorities, then the chosen of priority is arbitrary.
%\end{itemize}

%Let us introduce the notion of sequential executions, which modelled the specification of concurrent libraries. We fix a (possibly infinite) set $\mathbb{D}$ of data values, a finite set $\mathbb{M}$ of methods, and an infinite set $\mathbb{O}$ of operation (identifiers). Given $m \in \mathbb{M}$, $x \in \mathbb{D}$, and $o \in \mathbb{O}$, a call action $\textit{call}_o (m,x)$ represents an invocation to method $m$ with argument $x$, while a return action $\textit{ret}_o (m,x)$ represents a response from method $m$ with return value $x$. \footnote{Call actions with more than one arguments are similarly defined.} Here $o$ is used to match return actions to their call actions: $\textit{cal}_o (m,x)$ matches $\textit{ret}_{o'} (m',x')$, if $o=o' \wedge m=m'$. A sequential execution is a sequence of call and return actions, while each call action is immediately followed by its matching return action. Let $\cdot$ be the concatenation of sequences. To ease the reading, given a sequential execution $e=\textit{call}_{o_1}(m_1,a_1) \cdot \textit{rm}_{o_1}(m_1,b_1) \cdot \ldots \cdot \textit{call}_{o_n}(m_n,a_n) \cdot \textit{rm}_{o_n}(m_n,b_n)$, when the context is clear, we can write it as $e=m_1(a_1,b_1) \cdot \ldots \cdot m_n(a_n,b_n)$. We write $m(a,b)$ as $m(a)$ or $m(b)$, when there is no return value or there is no argument, respectively.

@@

We define two properties which are %satisfied by priority queue implementations and which are
important for our results: (1) \emph{data independence}~\cite{conf/popl/Wolper86,conf/tacas/AbdullaHHJR13} states that priority queue behaviors do not depend on the actual values which are added to the queue, and (2) \emph{closure under projection}~\cite{DBLP:conf/icalp/BouajjaniEEH15} states that executions remain valid by removing all the operations adding or removing certain values.
%remove operations can return the same values even if certain values are projected out from the priority queue state.
%no matter how many other different values are in the queue, provided that they don't have more important priorities.

%Data-independence \cite{Wolper:1986} can be used to effectively handle unbounded data domain. In this paper, we slightly modify the notion of data-independence in \cite{Wolper:1986} and propose data-independence for priority queues. Let $\_$ denote an element, of which the value is irrelevant.
An execution $e$ is \emph{data-differentiated} if every value is added at most once, i.e., for each $d \in \mathbb{D}$, $e$ contains at most one action $\textit{call}_o(\textit{put},d,p)$ with $o\in\mathbb{O}$ and $p\in \mathbb{P}$. Note that this property concerns only values, a data-differentiated execution $e$ may contain more than one value with the same priority. The subset of data-differentiated executions of a set of executions $E$ is denoted by $E_{\neq}$.

@@

%\vspace{-6pt}
\begin{definition}\label{def:priority-value data-independence}
A set of executions $E$ is \emph{data independent} iff
\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] for all $e \in E$, there exists $e' \in E_{\neq}$ and a renaming function $r$, such that $e=r(e')$,

@@

Besides data independence, the sequential executions of the priority queue satisfy the following closure property: an execution remains valid when removing all the operations with an argument in some set of values $D \subseteq \mathbb{D}$ and any $\textit{rm}(\textit{empty})$ operation (since they are read-only and they don't affect the queue's state).
To distinguish between different $\textit{rm}(\textit{empty})$ operations while simplifying the technical exposition, we assume that they receive as argument a value, i.e., call actions are of the form $\textit{call}_o(\textit{rm},\textit{empty},a)$ for some $a\in \mathbb{D}$. We will make explicit this argument only when needed in our technical development. The projection $e \vert D$ of an execution $e$ to a set of values $D \subseteq \mathbb{D}$ is obtained from $e$ by erasing all the call/return actions with an argument not in $D$. We write $e \setminus x$ for the projection $e \vert_{ \mathbb{D} \setminus \{ x \} }$. Let $\textit{proj}(e)$ be the set of all projections of $e$ to a set of values $D \subseteq \mathbb{D}$.
%{\color {red}The following lemma states that $\seqPQ$ is closed under projection.}

@@

%\begin{lemma}\label{lem:closure_proj}
%$\seqPQ$ is closed under projection, i.e., $\textit{proj}(e)\subseteq \seqPQ$ for each $e\in \seqPQ$.
%\end{lemma}
\begin{lemma}
\label{lem:closure_proj}
$\seqPQ$ is closed under projection, i.e., $\textit{proj}(e)\subseteq \seqPQ$ for each $e\in \seqPQ$.
\end{lemma}

@@

%A (concurrent) execution $e$ is a sequence of call and return actions which satisfy a well-formedness property: every return has a matching call action before it in $e$, and an operation $o$ can be used only twice in $e$, once in a call action, and once in a return action. The definition of data-differentiated, renaming function, data-independence and projection extends to (concurrent) executions. For instance, an execution is data-differentiated if, for all $d \in \mathbb{D}$, there is at most one $\textit{cal}_{\_}(\textit{put},d,\_)$.
%
%An implementation $\mathcal{I}$ is a set of concurrent executions. Implementations represent libraries whose methods are called by external programs. In the remainder of this work, we consider only completed executions, where each call action has a corresponding return action. This simplification is sound when implementation methods can always make progress in isolation \cite{Henzinger:2013}: formally, for any execution $e$ with pending operations, there exists an execution $e'$ obtained by extending $e$ only with the return actions of the pending operations of $e$. Intuitively this means that methods can always return without any help from outside threads, avoiding deadlock.
We recall the notion of \emph{linearizability}~\cite{journals/toplas/HerlihyW90} which is the \emph{de facto} standard correctness condition for concurrent data structures.
Given an execution $e$, the happen-before relation $<_{\textit{hb}}$ between operations~\footnote{In general, we refer to operations using their identifiers.} is defined as follows: $o_1 <_{\textit{hb}} o_2$, if the return action of $o_1$ occurs before the call action of $o_2$ in $e$. The happens-before relation is an interval order \cite{DBLP:conf/popl/BouajjaniEEH15}: for distinct $o_1,o_2,o_3,o_4$, if $o_1 <_{\textit{hb}} o_2$ and $o_3 <_{\textit{hb}} o_4$, then either $o_1 <_{\textit{hb}} o_4$, or $o_3 <_{\textit{hb}} o_2$. Intuitively, this comes from the fact that concurrent threads share a notion of global time.

Given a (concurrent) execution $e$ and a sequential execution $s$, we say that $e$ is linearizable w.r.t $s$, denoted $e \sqsubseteq s$, if there is a bijection $f: O_1 \rightarrow O_2$, where $O_1$ and $O_2$ are the set of operations of $e$ and $s$, respectively, such that (1) the call and return actions with identifier $o$ and $f(o)$, respectively, are the same
%{\color{blue}$o$ and $f(o)$ have the same call and return actions}{\color {red}$o$ and $f(o)$ is the same operation}\footnote{},
and (2) if $o_1 <_{\textit{hb}} o_2$, then $f(o_1) <_{\textit{hb}} f(o_2)$.
A (concurrent) execution $e$ is linearizable w.r.t. a set $S$ of sequential executions, denoted $e \sqsubseteq S$, if there exists $s \in S$ such that $e \sqsubseteq s$. A set of concurrent executions $E$ is linearizable w.r.t. $S$, denoted $E \sqsubseteq S$, if $e \sqsubseteq S$ for all $e \in E$.

The following lemma states that by data-independence, it is enough to consider only data-differentiated executions when checking linearizability (see Appendix \ref{sec:appendix proofs in section priority queue and data-independence}).
%This is similar to that in \cite{conf/tacas/AbdullaHHJR13,DBLP:conf/icalp/BouajjaniEEH15}, where they use the notion of data-independence in \cite{conf/popl/Wolper86}.
Section~\ref{sec:checking inclusion by recursive procedure} will focus on characterizing linearizability for data-differentiated executions. %The proof of this lemma can be found in Appendix \ref{sec:appendix in section data-independence of EPQ}.

>>>>

We consider priority queues whose interface contains two methods $\textit{put}$ and $\textit{rm}$ for adding and respectively, removing a value. Each value is assigned with a priority when being added to the data structure (by calling $\textit{put}$) and the remove method $\textit{rm}$ removes a value with a minimal priority. For generality, we assume that the set of priorities is partially-ordered. Incomparable priorities can be removed in any order. When multiple values are assigned with the same priority, $\textit{rm}$ returns the least recent value. Also, when the set of values stored in the priority queue is empty, $\textit{rm}$ returns the distinguished value $\textit{empty}$. In this section, we formalize (concurrent) executions and implementations, introduce a set of properties satisfied by all the implementations we are aware of, and recall the standard correctness criterion for concurrent implementations of ADTs known as \emph{linearizability}~\cite{journals/toplas/HerlihyW90}.

@@

We define $\seqPQ$, the set of sequential priority queue executions, semantically via a labelled transition system (LTS, for short). An LTS is a tuple $A=(Q,\Sigma,\rightarrow,q_0)$, where $Q$ is a set of states, $\Sigma$ is an alphabet of transition labels, $\rightarrow\subseteq Q\times\Sigma\times Q$ is a transition relation, and $q_0$ is the initial state. We model the priority queue as an LTS $\textit{PQ}$ where states are mappings associating priorities in $\mathbb{P}$ with sequences of values in $\mathbb{D}$, representing a snapshot of the priority queue (for each priority, the values are ordered as they were inserted), and the transition labels are $\textit{put}(a,p)$ and $\textit{rm}(a)$. Each transition modifies the state as expected. For example, $q_1 \xrightarrow{\textit{rm}(\textit{empty})} q_2$ if $q_1 = q_2$, and $q_1$ and $q_2$ map each priority to the empty sequence $\epsilon$. Then, $\seqPQ$ is the set of traces (words) accepted by $\textit{PQ}$. The detailed definition of $\textit{PQ}$ can be found in Appendix \ref{sec:appendix definition of seqPQ and proof of Lemma EQP rules and semantics}.

An implementation $\mathcal{I}$ is a set of executions. Implementations represent libraries whose methods are called by external programs. In the remainder of this work, we consider only \emph{completed} executions, where each call action has a corresponding return action. This simplification is sound when the method invocations can always make progress in isolation.

@@

We define two properties which are
important for our results: (1) \emph{data independence}~\cite{conf/popl/Wolper86,conf/tacas/AbdullaHHJR13} states that priority queue behaviors do not depend on the actual values which are added to the queue, and (2) \emph{closure under projection}~\cite{DBLP:conf/icalp/BouajjaniEEH15} states that executions remain valid by removing all the operations adding or removing certain values.

An execution $e$ is \emph{data-differentiated} if every value is added at most once, i.e., for each $d \in \mathbb{D}$, $e$ contains at most one action $\textit{call}_o(\textit{put},d,p)$ with $o\in\mathbb{O}$ and $p\in \mathbb{P}$. Note that this property concerns only values, a data-differentiated execution $e$ may contain more than one value with the same priority. The subset of data-differentiated executions of a set of executions $E$ is denoted by $E_{\neq}$.

@@

\begin{definition}\label{def:priority-value data-independence}
A set of executions $E$ is \emph{data independent} iff
\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] for all $e \in E$, there exists $e' \in E_{\neq}$ and a renaming function $r$, such that $e=r(e')$,

@@

Besides data independence, the sequential executions of the priority queue satisfy the following closure property: an execution remains valid when removing all the operations with an argument in some set of values $D \subseteq \mathbb{D}$ and any $\textit{rm}(\textit{empty})$ operation (since they are read-only and they don't affect the queue's state).
To distinguish between different $\textit{rm}(\textit{empty})$ operations while simplifying the technical exposition, we assume that they receive as argument a value, i.e., call actions are of the form $\textit{call}_o(\textit{rm},\textit{empty},a)$ for some $a\in \mathbb{D}$. We will make explicit this argument only when needed in our technical development. The projection $e \vert D$ of an execution $e$ to a set of values $D \subseteq \mathbb{D}$ is obtained from $e$ by erasing all the call/return actions with an argument not in $D$. We write $e \setminus x$ for the projection $e \vert_{ \mathbb{D} \setminus \{ x \} }$. Let $\textit{proj}(e)$ be the set of all projections of $e$ to a set of values $D \subseteq \mathbb{D}$.

@@

\begin{lemma}
\label{lem:closure_proj}
$\seqPQ$ is closed under projection, i.e., $\textit{proj}(e)\subseteq \seqPQ$ for each $e\in \seqPQ$.
\end{lemma}

@@

We recall the notion of \emph{linearizability}~\cite{journals/toplas/HerlihyW90} which is the \emph{de facto} standard correctness condition for concurrent data structures.
Given an execution $e$, the happen-before relation $<_{\textit{hb}}$ between operations~\footnote{In general, we refer to operations using their identifiers.} is defined as follows: $o_1 <_{\textit{hb}} o_2$, if the return action of $o_1$ occurs before the call action of $o_2$ in $e$. The happens-before relation is an interval order \cite{DBLP:conf/popl/BouajjaniEEH15}: for distinct $o_1,o_2,o_3,o_4$, if $o_1 <_{\textit{hb}} o_2$ and $o_3 <_{\textit{hb}} o_4$, then either $o_1 <_{\textit{hb}} o_4$, or $o_3 <_{\textit{hb}} o_2$. Intuitively, this comes from the fact that concurrent threads share a notion of global time.

Given a (concurrent) execution $e$ and a sequential execution $s$, we say that $e$ is linearizable w.r.t $s$, denoted $e \sqsubseteq s$, if there is a bijection $f: O_1 \rightarrow O_2$, where $O_1$ and $O_2$ are the set of operations of $e$ and $s$, respectively, such that (1) the call and return actions with identifier $o$ and $f(o)$, respectively, are the same
and (2) if $o_1 <_{\textit{hb}} o_2$, then $f(o_1) <_{\textit{hb}} f(o_2)$.
A (concurrent) execution $e$ is linearizable w.r.t. a set $S$ of sequential executions, denoted $e \sqsubseteq S$, if there exists $s \in S$ such that $e \sqsubseteq s$. A set of concurrent executions $E$ is linearizable w.r.t. $S$, denoted $E \sqsubseteq S$, if $e \sqsubseteq S$ for all $e \in E$.

The following lemma states that by data-independence, it is enough to consider only data-differentiated executions when checking linearizability (see Appendix \ref{sec:appendix proofs in section priority queue and data-independence}).
Section~\ref{sec:checking inclusion by recursive procedure} will focus on characterizing linearizability for data-differentiated executions.

### 2017-07-13 related.tex
arxiv

Existing automated methods for proving linearizability of a concurrent object
implementation are also based on reductions to safety
verification, e.g.,~\cite{conf/tacas/AbdullaHHJR13, conf/concur/HenzingerSV13,
conf/cav/Vafeiadis10}. The approach in~\cite{conf/cav/Vafeiadis10} considers
implementations where 
operations' \emph{linearization points}
are 
% fixed to particular source-code locations.
manually specified.
Essentially, this approach instruments the
implementation with ghost variables simulating the ADT specification at
linearization points. This approach is incomplete since not all implementations
have fixed linearization points. Aspect-oriented
proofs~\cite{conf/concur/HenzingerSV13} reduce linearizability to the
verification of four simpler safety properties. However, this approach has only
been applied to queues, and has not produced a fully automated
and complete proof technique. The work in~\cite{Dodds:2015:SCT:2676726.2676963} proves 
linearizability of stack implementations with an automated proof assistant. 
Their approach does not lead to full automation however, e.g.,~by reduction to 
safety verification.

@@

%
%these results require establishing key results which are specific to priority queues and which are not implied by those for stacks and queues (say that more details are given in the related work)
%
%\noindent The most related work of our paper is ~\cite{DBLP:conf/icalp/BouajjaniEEH15}. Compared to ~\cite{DBLP:conf/icalp/BouajjaniEEH15}, our work are specific to priority queue and thus have points not occur in queues and stacks. Due to the partial-order priorities, to prove the correctness of recursive procedure, we need to alter the positions of linearization points of incomparable items while keeps the linearization points of comparable items unchanged. {\color {blue}In the proof process we use the notion of left-right constraint, which is inspired by left-right constraint of queue \cite{Bouajjani:2015}.} For checking violations of one local property, we just need to find a value, such that at every possible time point to locate its remove operation, there is already values with smaller priorities in priority queue. For checking violations of another local property, we just need to find two values with maximal priority, such that every possible time point to locate remove operation of one value is disabled by remove operation of another value. Such phenomenons are unique for priority queue.

>>>>

Existing automated methods for proving linearizability of a concurrent object
 are also based on reductions to safety
verification, e.g.,~\cite{conf/tacas/AbdullaHHJR13, conf/concur/HenzingerSV13,
conf/cav/Vafeiadis10}. The approach in~\cite{conf/cav/Vafeiadis10} considers
implementations where
operations' \emph{linearization points}
are
manually specified.
Essentially, this approach instruments the
implementation with ghost variables simulating the ADT specification at
linearization points. This approach is incomplete since not all implementations
have fixed linearization points. Aspect-oriented
proofs~\cite{conf/concur/HenzingerSV13} reduce linearizability to the
verification of four simpler safety properties. This approach has only
been applied to queues, and has not produced a fully automated
and complete proof technique. The work in~\cite{Dodds:2015:SCT:2676726.2676963} proves
linearizability of stack implementations with an automated proof assistant.
Their approach does not lead to full automation however, e.g.,~by reduction to
safety verification.

@@
